Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Fixed some stuff

@@

<{:const:gpuAlloc}>
<{:const:gpuRelease}>

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the calls to <{release_one()}> in
\autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to <{M}> threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M = 1} is an ordinary lock.
You should define a constant <{M}> and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than <{M}> threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to <{M}> threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method <{push(item}>)
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are <{N}> GPUs identified by the numbers
1 through <{N}>.  Method <{gpuAlloc()}> returns the identifier
of an available GPU, blocking if there is currently no GPU available.
Method <{gpuRelease(gpu}>) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a <{downgrade}> method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.  Why is this problematic?)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

Note that <{wait}> is always invoked within a <{while}>
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a <{while}> loop
around any invocation of <{wait}> containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where <{wait}> may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace <{wait}>
by <{release}> followed by <{acquirire}>.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

<{:module:qsort}>
<{:const:Qsort}>
<{:const:swap}>
<{:const:partition}>
<{:const:sortrange}>
<{:const:sort}>
<{:const:main}>

\section*{Exercises}
\begin{problems}
\item \label{ex:bbmesa} Implement a solution to the bounded buffer problem using Mesa condition
variables.
\item Implement a ``try lock'' module using Mesa condition variables
(see also \autoref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item <{tl = TryLock()}> \# \emph{create a try lock}
\item <{acquire(?tl}>) \# \emph{acquire a try lock}
\item <{tryAcquire(?tl}>) \# \emph{attempt to acquire a try lock}
\item <{release(?tl}>) \# \emph{release a try lock}
\end{enumerate}
\noindent
<{tryAcquire}> should not wait.
Instead it should return <{True}> if the lock was successfully
acquired and <{False}> if the lock was not available.
% \item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a thread is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
It returns a set of $n$ GPU identifiers.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A thread does not have to return all the GPUs it allocated at once.
(You may want to try implementing this with Split Binary Semaphores.  It is not as easy.)
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\item \label{ex:qsort}
Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and \autoref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in shared variable <{textqs.arr}>.
Another shared variable, <{testqs.todo}>, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods <{swap}>, <{partition}>,
or <{sortrange}> for this.  Create <{NWORKERS}> ``worker threads''
that should replace the <{qsort}> code.
Each worker loops until <{todo}>
is empty and sorts the ranges that it finds until then.  The <{main}>
thread needs to wait until all workers are done.
\end{problems}

>>>>

<{:const:tryAcquire}>

@@

<{:const:gpuAlloc}>
<{:const:gpuRelease}>
<{:const:acquire_bound_lock}>
<{:const:release_bound_lock}>
<{:const:nb_enter}>
<{:const:nb_leave}>
<{:const:sb_enter}>
<{:const:sb_leave}>

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the calls to <{release_one()}> in
\autoref{fig:RWsplitsema} can be replaced by simply releasing <{mutex}>.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to <{M}> threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M = 1} is an ordinary lock.
You should define a constant <{M}> and two methods:
<{acquire_bound_lock()}> and <{release_bound_lock()}>.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than <{M}> threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to <{M}> threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method <{push(item}>)
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are <{N}> GPUs identified by the numbers
1 through <{N}>.  Method <{gpuAlloc()}> returns the identifier
of an available GPU, blocking if there is currently no GPU available.
Method <{gpuRelease(gpu}>) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a <{downgrade}> method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.  Why is this problematic?)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
<{nb_enter()}> that a car must invoke before going northbound on
the bridge and <{nb_leave()}> that the car must invoke after leaving
the bridge.  Similarly write <{sb_enter()}> and <{sb_leave()}>
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

Note that <{wait}> is always invoked within a <{while}>
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a <{while}> loop
around any invocation of <{wait}> containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where <{wait}> may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace <{wait}>
by <{release}> followed by <{acquire}>.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

<{:const:TryLock}>
<{:module:qsort}>
<{:const:Qsort}>
<{:const:swap}>
<{:const:partition}>
<{:const:sortrange}>
<{:const:sort}>
<{:const:main}>

\section*{Exercises}
\begin{problems}
\item \label{ex:bbmesa} Implement a solution to the bounded buffer problem using Mesa condition
variables.
\item Implement a ``try lock'' module using Mesa condition variables
(see also \autoref{ex:trylock}).  It should
have the following API:
\begin{enumerate}
\item <{tl = TryLock()}> \# \emph{create a try lock}
\item <{acquire(?tl}>) \# \emph{acquire a try lock}
\item <{tryAcquire(?tl}>) \# \emph{attempt to acquire a try lock}
\item <{release(?tl}>) \# \emph{release a try lock}
\end{enumerate}
\noindent
<{tryAcquire}> should not wait.
Instead it should return <{True}> if the lock was successfully
acquired and <{False}> if the lock was not available.
% \item Implement semaphores using Mesa condition variables.
\item Write a new version of the GPU allocator in \autoref{ex:gpu}
using Mesa condition variables.
In this version,
a thread is allowed to allocate a set of GPUs and release a set of GPUs that it
has allocated.  Method \texttt{gpuAllocSet(n)} should block until $n$ GPUs are
available, but it should grant them as soon as they are available.
It returns a set of $n$ GPU identifiers.
Method \texttt{gpuReleaseSet(s)} takes a set of GPU identifiers as argument.
A thread does not have to return all the GPUs it allocated at once.
(You may want to try implementing this with Split Binary Semaphores.  It is not as easy.)
\item The specification in the previous question makes the solution unfair.
Explain why this is so.  Then change the specification and the solution so that
it is fair.
\item \label{ex:qsort}
Bonus problem: \autoref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and \autoref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in shared variable <{testqs.arr}>.
Another shared variable, <{testqs.todo}>, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods <{swap}>, <{partition}>,
or <{sortrange}> for this.  Create <{NWORKERS}> ``worker threads''
that should replace the <{qsort}> code.
Each worker loops until <{todo}>
is empty and sorts the ranges that it finds until then.  The <{main}>
thread needs to wait until all workers are done.
\end{problems}
